% !TeX root = ../../thesis.tex
\chapter{Curriculum Vitae}\label{ch:curriculum}

Jelle Vermandere was born on June 26, 1995. In 2020, he obtained his master’s degree in Architectural Engineering at UGent, graduating with distinction. 
His thesis focussed on creating a website and database for interactive viewing of 3D models withing exhibitions.

After his studies, Jelle was a self-employed freelancer, creating interactive architectural renderings and video games for various clients.

In July 2021, he started his research at the Faculty of engineering technology at the KU Leuven under the supervision of Prof. Vergauwen and Prof. Bassier. 
A year later, he secured an FWO strategic basic research grand, allowing him to focus his main research which centers around innovating the built environment through Reality Modelling.
The goal of this research is to create dynamic and complete 3D models of indoor environments captured using 3D scan-devices.

Jelle has a wide research interest, focussing around the combination of real and digital techniques. 
Outside his research, his hobbies led him to mastering the full real-to-digital-to-real workflow using laser scanning, 3D modeling and 3D printing, he can create almost anything.
He documents his projects on his YouTube channel where he created a community of over 90.000 subscribes accumulating over 7 million views.

Jelle also likes to practice a wide range of sports including: running, biking, swimming and climbing. 
After all the hours behind his computer, the physical exercise is a very welcome distraction.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Keep the following \cleardoublepage at the end of this file, 
% otherwise \includeonly includes empty pages.
\cleardoublepage

% vim: tw=70 nocindent expandtab foldmethod=marker foldmarker={{{}{,}{}}}
