% !TeX root = ../../thesis.tex
\chapter{Data Acquisition}
\label{ch:3_Data_Acquisition}

% ── New glossary entries ─────────────────────────────────────────────────────
\newglossaryentry{TLS}{name={TLS}, description={Terrestrial Laser Scanner}}
\newglossaryentry{iMMS}{name={iMMS}, description={Indoor Mobile Mapping System}}
\newglossaryentry{ToF}{name={ToF}, description={Time-of-Flight}}
\newglossaryentry{LOA}{name={LOA}, description={Level of Accuracy}}
\newglossaryentry{SLAM}{name={SLAM}, description={Simultaneous Localisation and Mapping}}
\newglossaryentry{TSDF}{name={TSDF}, description={Truncated Signed Distance Field}}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:3_intro}

The pipeline introduced in Chapter~\ref{ch:background} requires data that satisfies several demanding conditions simultaneously. The input must consist of partially scanned indoor single-room scenes with accompanying colour information, captured from both industrial and residential environments. To support the alignment and updating stage (Chapter~\ref{ch:4_Alignment_Updating}), multi-temporal and multi-sensory captures of the same environment are needed. The object detection stage (Chapter~\ref{ch:5_Scene_Segmentation}) requires ground-truth annotations for detectable objects, including bounding boxes. The scene completion stage (Chapter~\ref{ch:6_Scene_Completion}) requires ground-truth geometry of the empty scene, while the object completion stage (Chapter~\ref{ch:7_Object_Completion}) requires both ground-truth complete object geometry and isolated partial scans of individual objects.

No single existing dataset satisfies all of these requirements. Publicly available scene-level datasets such as ScanNet~\cite{dai_scannet_2017} and Matterport3D~\cite{chang_matterport3d_2017} provide real scans with instance segmentation annotations, but lack ground-truth geometry for occluded regions. Synthetic scene-level datasets offer controlled conditions but typically lack the noise characteristics and occlusion patterns of real scanning devices. At the object level, large synthetic repositories such as ShapeNet~\cite{chang_shapenet_2015} provide a wide variety of complete models, but these are not scanned data and often contain self-intersecting geometry. Real object-level datasets such as Redwood~\cite{choi_large_2016} offer high-quality reconstructions but cover a limited number of objects.

To address these gaps, this chapter describes the data acquisition strategy adopted in this thesis. Three complementary data sources are combined. First, publicly available datasets are leveraged for specific pipeline stages where they provide sufficient coverage (Section~\ref{sec:3_existing}). Second, real-world multi-temporal and multi-sensory captures are recorded at the university campus, providing heterogeneous data from different scanning technologies for the alignment and updating stage (Section~\ref{sec:3_realworld}). Third, a novel synthetic dataset generation framework is developed, consisting of a virtual scanner and a procedural room layout generator, to produce the V-Scan dataset. This dataset provides scene-level captures with and without objects, ground-truth geometry for all detectable objects along with their bounding boxes and isolated scans, and multi-temporal and multi-sensory variations, all under controlled conditions that replicate real-world scanning characteristics (Section~\ref{sec:3_synthetic}). Section~\ref{sec:3_comparison} presents a comparative overview of all datasets used in this thesis, and Section~\ref{sec:3_conclusion} concludes the chapter.

% ─────────────────────────────────────────────────────────────────────────────
\section{Existing Datasets}
\label{sec:3_existing}

Several publicly available datasets have become standard benchmarks for 3D scene understanding and object reconstruction tasks. This section reviews the datasets used in this thesis and discusses their characteristics and limitations with respect to the requirements defined above.

\subsection{Scene-Level Datasets}
\label{sec:3_scene_datasets}

\paragraph{ScanNet.}
ScanNet~\cite{dai_scannet_2017} is an RGB-D video dataset containing 2.5 million views across more than 1500 scans, annotated with 3D camera poses, surface reconstructions, and instance-level semantic segmentations. Scenes are captured using commodity RGB-D sensors and reconstructed using BundleFusion. The dataset covers a wide variety of indoor environments including offices, bedrooms, kitchens, and living rooms. ScanNet provides per-vertex semantic and instance labels but does not include ground-truth geometry for occluded regions, limiting its use for quantitative evaluation of completion methods. The dataset is used in this thesis primarily for qualitative evaluation and to assess generalisation of detection methods.

\paragraph{ScanNet++.}
ScanNet++~\cite{yeshwanth_scannet_2023} extends the ScanNet paradigm with significantly higher fidelity. The dataset currently contains over 1000 scenes, each captured with a high-end laser scanner at sub-millimetre resolution, registered 33-megapixel DSLR images, and commodity RGB-D streams from an iPhone. Scene reconstructions are annotated with an open vocabulary of semantics. The combination of high-quality laser scans and commodity-level captures in the same scenes makes ScanNet++ particularly relevant for evaluating methods that operate on heterogeneous data. However, as with ScanNet, no ground-truth completion data is available.

\paragraph{Matterport3D.}
Matterport3D~\cite{chang_matterport3d_2017} provides 10,800 panoramic views from 194,400 RGB-D images across 90 building-scale scenes, captured with the Matterport Pro 3D Camera. The dataset includes textured surface reconstructions, globally aligned camera poses, and both 2D and 3D semantic segmentations at the instance level. Scenes span entire buildings, covering multiple rooms across several floors, and include a diverse set of residential environments. The building-scale coverage and panoramic image availability make Matterport3D well suited for evaluating scene-level methods that operate on panoramic imagery. The dataset is used in this thesis for qualitative evaluation of the scene completion and object completion stages.

\paragraph{ARKitScenes.}
ARKitScenes~\cite{baruch_arkitscenes_2022} provides a large collection of indoor scenes captured with Apple ARKit on iPad Pro devices equipped with LiDAR sensors. The dataset includes RGB-D video streams, camera poses, and 3D annotations for oriented bounding boxes of common indoor objects. ARKitScenes represents the class of commodity-level capture devices that the alignment stage of this pipeline must accommodate.

\subsection{Object-Level Datasets}
\label{sec:3_object_datasets}

\paragraph{ShapeNet.}
ShapeNet~\cite{chang_shapenet_2015} is a large-scale repository of 3D CAD models organised according to the WordNet taxonomy, containing over 51,000 unique models across 55 common object categories. The models are clean, watertight, and consistently oriented, making them well suited as training data for object completion networks. However, as synthetic CAD models, they differ substantially from scanned data in terms of surface noise, scan artefacts, and occlusion patterns. ShapeNet is used in this thesis as training data for the object completion networks described in Chapter~\ref{ch:7_Object_Completion}.

\paragraph{Redwood.}
The Redwood 3D Scan Dataset~\cite{choi_large_2016} provides high-quality 3D reconstructions of real-world objects captured with consumer-grade RGB-D sensors. The dataset includes both individual object scans and full scene reconstructions. While the object reconstructions are detailed and realistic, the dataset covers a relatively limited number of objects compared to synthetic alternatives, restricting its use as a comprehensive training set.

% ─────────────────────────────────────────────────────────────────────────────
\section{Real-World Data Capture}
\label{sec:3_realworld}

\subsection{Introduction}

The first set of manually scanned data used in this thesis consists of real-world captures performed at the KU~Leuven Technology Campus in Ghent, Belgium. These captures serve two primary purposes: providing multi-temporal and multi-sensory data for evaluating the alignment and updating pipeline (Chapter~\ref{ch:4_Alignment_Updating}), and providing real indoor scenes with diverse object configurations for qualitative evaluation of the detection and completion stages. The captures cover a range of industrial laboratory spaces and residential-style environments, using a combination of high-end terrestrial laser scanners and consumer-grade \gls{XR} devices.

\subsection{Sensing Technologies}
\label{sec:3_sensors}

The sensing technologies used in this thesis span a wide range of accuracy, density, and operational characteristics. Their properties directly influence the nature of the captured data and the challenges that each pipeline stage must address.

\paragraph{Terrestrial Laser Scanners.}
Static \gls{TLS} systems such as the Leica ScanStation P30 provide the highest measurement accuracy, typically achieving single-point accuracies below 5\,mm at ranges up to 50\,m~\cite{bassier_standalone_2016}. These systems capture structured point grids of 10 to 40 million points per setup, along with panoramic imagery. However, the static nature of the measurement introduces significant occlusions at ground level and behind objects, as only surfaces visible from the scanner position are recorded. Consecutive setups must be registered, typically through a combination of target-based and cloud-to-cloud methods, with sub-millimetre error propagation achievable across setups~\cite{geyter_point_2022}.

\paragraph{Indoor Mobile Mapping Systems.}
Indoor mobile mapping systems (\gls{iMMS}) such as the NavVis VLX combine LiDAR sensors with camera-based \gls{SLAM} algorithms to achieve continuous data acquisition while the operator moves through the environment. These systems produce unordered point clouds with uneven point distributions, but their continuous trajectory results in significantly better spatial coverage than static scanners, particularly in cluttered environments. Global accuracies of LOA20 ($2\sigma \leq 0.05$\,m) to LOA30 ($2\sigma \leq 0.015$\,m) are achievable when supported by control measurements~\cite{geyter_point_2022, vermandere_two-step_2022}. Data acquisition is typically two to three times faster than with static TLS.

\paragraph{XR Devices.}
Extended Reality devices such as the Microsoft HoloLens~2 represent the lower end of the capture spectrum. These head-mounted devices are equipped with a 1\,MP \gls{ToF} sensor and four grey-scale tracking cameras, producing real-time polygonal meshes at a default spatial resolution of approximately 0.08\,m$^3$~\cite{geyter_point_2022}. While their accuracy is limited, with single-point accuracies of approximately 1\,cm/m and a maximum useful range of 3--4\,m, they offer unparalleled mobility and can capture regions inaccessible to larger systems. The accompanying RGB camera provides localised imagery that can be used for 2D-based alignment methods.

\paragraph{Smartphone LiDAR.}
Recent smartphones equipped with LiDAR sensors, such as iPhone models with the Apple LiDAR Scanner, provide another commodity-level capture modality. These devices produce dense point clouds at close range with moderate accuracy, and their widespread availability makes them relevant for incremental updating scenarios where dedicated scanning equipment is not practical.

\subsection{Capture Methodology}
\label{sec:3_capture_method}

Each sensing technology was deployed according to its intended use case. For TLS captures, one or two scanner setups were placed per room, positioning the scanner to maximise coverage of the structural elements while accepting the inherent ground-level occlusions caused by furniture and equipment. For iMMS captures, the operator walked through the environment following a natural trajectory, ensuring loop closures where possible to minimise drift accumulation. For XR device captures, the operator walked around the room attempting to capture all accessible surfaces, including regions underneath and behind furniture that are typically occluded in TLS data.

This deliberate combination of sensors and deployment strategies produces complementary datasets: TLS data provides high-accuracy structural geometry with systematic occlusions, iMMS data provides high-coverage reconstructions with moderate accuracy, and XR data provides fast, accessible captures with reduced accuracy but improved coverage in cluttered areas.

\subsection{Captured Scenes}
\label{sec:3_captured_scenes}

Multiple indoor environments were captured across the KU~Leuven campus. These include industrial laboratory spaces used for concrete processing, road construction research, and building physics experiments, as well as a residential-style living lab environment. Each environment was captured at multiple time points and with multiple sensors, producing the multi-temporal and multi-sensory datasets required for the alignment experiments in Chapter~\ref{ch:4_Alignment_Updating}. Between capture epochs, the environments were in active use, resulting in natural variations in object placement and configuration.

The captured data from these campaigns has been previously reported and analysed in~\cite{geyter_point_2022, vermandere_two-step_2022, vermandere_automatic_2022}, where the sensor capacity, alignment accuracy, and data processing characteristics were evaluated in detail.

\subsection{Comparison of Sensing Technologies}
\label{sec:3_sensor_comparison}

The sensor comparison studies reported in~\cite{geyter_point_2022} provide quantitative evidence for the complementary nature of the different capture technologies. The TLS achieves the highest point accuracy (LOA40) but the lowest spatial coverage (average 41\% across structure classes), primarily due to the limited number of setups and the resulting systematic occlusions. The backpack-based iMMS (NavVis VLX) achieves the highest coverage (average 60\%) while maintaining LOA20 to LOA30 accuracy, making it the most balanced system for general documentation. The HoloLens~2 achieves coverage comparable to the iMMS in accessible areas but is limited by its short range, resulting in poor coverage of ceilings, high walls, and distant surfaces. Its point accuracy is restricted to LOA10 to LOA20.

An important finding from~\cite{geyter_point_2022} is that the differences in point cloud characteristics between sensors have a limited impact on downstream semantic segmentation performance. This sensor invariance opens the door to combining training data from multiple sources and to using synthetic data for training deep learning models, which directly motivates the synthetic dataset generation described in the following section.

% ─────────────────────────────────────────────────────────────────────────────
\section{Synthetic Data Creation}
\label{sec:3_synthetic}

\subsection{Introduction}

While the real-world captures described in the previous section provide heterogeneous multi-sensory data for alignment evaluation, they cannot provide the ground-truth geometry required for quantitative evaluation of scene and object completion. The occluded surfaces that completion methods must reconstruct are by definition not observed in any real scan. This motivates the development of a synthetic dataset generation framework that produces partial scans paired with complete ground-truth geometry under controlled conditions that replicate real-world scanning characteristics.

The framework consists of two major components: a virtual scanner that simulates the acquisition process of real scanning devices (Section~\ref{sec:3_virtual_scanner}), and a procedural room layout generator that produces diverse furnished indoor environments (Section~\ref{sec:3_room_generator}). Together, these components enable the automated generation of the V-Scan dataset (Section~\ref{sec:3_vscan}).

\subsection{Related Work}
\label{sec:3_synthetic_related}

Synthetic datasets have been widely used in 3D computer vision to overcome the limitations of real-world data collection. ShapeNet~\cite{chang_shapenet_2015} provides a large collection of clean CAD models but does not simulate scanning artefacts. Structured3D~\cite{vedaldi_structured3d_2020} generates photorealistic indoor scenes with ground-truth annotations but does not simulate point cloud acquisition. Virtual scanning approaches have been explored in the context of autonomous driving, where LiDAR simulators generate realistic point clouds from virtual environments. In the indoor domain, our previous work~\cite{de_geyter_automated_2022} demonstrated that synthetic training data generated from virtual scanning can be combined with real data to improve semantic segmentation performance. However, existing virtual scanners typically focus on a single scanner type and do not provide the multi-sensory simulation required for this thesis.

\subsection{Virtual Scanner}
\label{sec:3_virtual_scanner}

The virtual scanner is implemented as a Unity component that simulates the acquisition process of different scanning technologies within a 3D scene. The scanner operates from a fixed position within the scene and produces a coloured point cloud along with a panoramic image, replicating the output of a terrestrial laser scanner or similar device.

\subsubsection{Scanner Configuration}

The scanner behaviour is controlled by a set of configurable parameters encapsulated in a \texttt{ScannerTypeScriptableObject}. These parameters define the characteristics of the simulated sensing device:

\begin{itemize}
    \item \textbf{Scan density}: the angular spacing between consecutive rays, expressed as the point spacing in mm at 10\,m distance. This parameter controls the total number of rays cast and thus the density of the resulting point cloud.
    \item \textbf{Scan range}: the maximum distance at which surfaces are detected, in metres.
    \item \textbf{Vertical scan range}: the total angular extent of the vertical field of view, in degrees. A value less than 360° simulates the blind spots typically found at the zenith and nadir of terrestrial laser scanners.
    \item \textbf{System noise}: the standard deviation of the range-independent measurement noise, in mm. This component models the intrinsic noise floor of the sensor electronics.
    \item \textbf{Distance noise}: the standard deviation of the range-dependent measurement noise, in mm per 10\,m. This component models the increase in measurement uncertainty with distance.
\end{itemize}

To replicate the characteristics of real scanning devices, a set of scanner type presets was created based on manufacturer specifications. Table~\ref{tab:scanner_presets} summarises the parameters of the simulated scanner types used in this thesis.

\begin{table}[h]
    \centering
    \caption[Virtual scanner presets]{Scanner type presets used in the virtual scanning framework, based on manufacturer specifications of real devices.}
    \label{tab:scanner_presets}
    \begin{tabular}{l c c c c c}
        \toprule
        \textbf{Scanner Type} & \textbf{Density} & \textbf{Range} & \textbf{Vert.\ Range} & \textbf{Sys.\ Noise} & \textbf{Dist.\ Noise} \\
         & (mm/10m) & (m) & (°) & (mm) & (mm/10m) \\
        \midrule
        Leica P30          & 12.5 & 100 & 290 & 1.0 & 1.0 \\
        Leica BLK360 Dense & 6.0  & 60  & 300 & 4.0 & 7.0 \\
        Leica BLK360 Fast  & 12.5 & 60  & 300 & 4.0 & 7.0 \\
        Matterport Pro2    & 25.0 & 9   & 260 & 3.0 & 5.0 \\
        Matterport Pro3    & 20.0 & 30  & 280 & 2.0 & 3.0 \\
        iPhone LiDAR       & 50.0 & 5   & 180 & 10.0 & 10.0 \\
        HoloLens 2         & 80.0 & 4   & 120 & 10.0 & 15.0 \\
        \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Geometric Scanning}

The scanning process operates by casting rays from the scanner position into the scene. The scan directions are computed to approximate the spherical scanning pattern of a terrestrial laser scanner. For a given angular density $\delta$, the number of horizontal points per disc and the number of vertical discs are determined by:

\begin{equation}
    n_{\text{horiz}} = \left\lceil \frac{2\pi}{\delta} \right\rceil, \qquad
    n_{\text{vert}} = \left\lceil \frac{\pi}{\delta} \right\rceil
\end{equation}

where $\delta = 2 \sin(\text{density} / 10000)$ converts the density parameter from mm at 10\,m to radians. Rays outside the configured vertical scan range are discarded to simulate the blind zones of real scanners.

For computational efficiency, the raycasting is parallelised using Unity's \texttt{RaycastCommand} batch processing within the C\# Job System. All rays are scheduled as a single batch and processed in parallel, significantly reducing computation time compared to sequential raycasting.

When a ray intersects a scene surface within the configured scan range, a scanned point is recorded with its 3D position, surface normal, and UV coordinates for subsequent colour assignment. If noise simulation is enabled, the measured distance is perturbed by Gaussian noise generated using the Box--Muller transform:

\begin{equation}
    d' = d + \mathcal{N}\left(0,\; \sigma_{\text{sys}} \cdot 10^{-3} + \sigma_{\text{dist}} \cdot 10^{-4} \cdot d\right)
\end{equation}

where $d$ is the true distance, $\sigma_{\text{sys}}$ is the system noise in mm, and $\sigma_{\text{dist}}$ is the distance noise in mm/10m.

\subsubsection{Image Capture and Colouring}

Colour information is captured through a spherical image render using a dedicated camera component (\texttt{SphericalImageCapture}). The camera renders an equirectangular panoramic image from the scanner position, which serves both as the panoramic image output and as the source for point cloud colouring.

Each scanned point is assigned a colour by sampling the panoramic image at its corresponding UV coordinates. This sampling is parallelised using a \texttt{IJobParallelFor} implementation that maps each point's spherical UV coordinates to pixel indices in the rendered texture.

\subsubsection{Object Isolation}

A key feature of the virtual scanner for dataset generation is the ability to isolate individual objects from the scene scan. Each object in the scene is annotated with a \texttt{CaptureObject} component that defines its bounding box and provides methods for isolating the points that fall within its volume from the full scene point cloud. The isolation is performed by filtering the scanned points against the object's oriented bounding box.

For each isolated object, the framework exports the partial point cloud (the scanned points within the bounding box), the ground-truth mesh (the complete mesh geometry of the object), and the oriented bounding box parameters. Optionally, a per-object voxel occupancy grid can be computed to characterise the occlusion pattern within the object's bounding box.

\subsubsection{Export Format}

The virtual scanner exports point clouds in a text-based format compatible with CloudCompare, with each line containing the position, normal, and colour of a scanned point (nine values: $x$, $y$, $z$, $n_x$, $n_y$, $n_z$, $r$, $g$, $b$). A header block records the capture timestamp and the scanner's transformation matrix. Ground-truth meshes are exported in PLY format, and bounding boxes and scan parameters are serialised as JSON files.

\subsection{Procedural Room Layout Generator}
\label{sec:3_room_generator}

The procedural room generator produces randomised indoor environments consisting of a rectangular room structure populated with furniture objects. The generator is implemented as a Unity component (\texttt{ProceduralRoomGenerator}) that creates rooms from a library of modular prefab elements defined in a \texttt{PrefabListScriptableObject}.

\subsubsection{Room Structure Generation}

The room structure is generated on a grid-based system. The floor dimensions are randomly sampled within configurable bounds, and the floor, walls, and ceiling are assembled from prefab tiles placed on the grid. During wall generation, each wall segment has a configurable probability of being replaced by a door or window prefab, introducing structural variation across generated rooms. A door is always placed at one wall position to ensure accessibility.

\subsubsection{Furniture Placement}

Furniture placement is divided into two categories: wall-mounted furniture and free-standing furniture. Wall-mounted furniture (such as shelves, paintings, or cabinets) is placed against available wall segments with a configurable probability. Free-standing furniture is placed within the interior of the room using a rejection sampling strategy. A placement zone is defined by insetting the room boundary by one grid cell, and objects are placed at random positions and orientations (quantised to 90° increments) within this zone. Each candidate placement is checked against all previously placed objects using axis-aligned bounding box intersection tests, and rejected if an overlap is detected. The maximum number of furniture items is controlled by a configurable density parameter that scales with the floor area.

The room structure and furniture are maintained as separate hierarchies, allowing the furniture to be regenerated independently of the room structure. This separation enables the creation of multi-temporal variations of the same room with different furniture configurations, which is essential for generating the scan variations described below.

\subsubsection{Seeded Randomisation}

All random decisions in the generation process are controlled by a single seed value, ensuring full reproducibility. When a fixed seed is specified, the same room layout and furniture configuration are produced deterministically. When no fixed seed is provided, the current system time is used, producing a unique configuration for each generation.

\subsection{Scan Management and Dataset Generation}
\label{sec:3_scan_management}

The virtual scanner and procedural room generator are coordinated by a \texttt{ScanManager} component that orchestrates the automated generation of complete dataset entries. For each generated room, the scan manager executes the following sequence:

\begin{enumerate}
    \item \textbf{Room generation}: A new room is procedurally generated, and all detectable objects are identified within the scan volume.
    \item \textbf{Full scene scan}: The scene is scanned with objects present, producing the main point cloud and panoramic image. Ground-truth meshes and bounding boxes are exported for all detected objects.
    \item \textbf{Object isolation}: Each detected object's points are isolated from the full scan, producing partial point clouds per object. Optionally, per-object occlusion voxel grids are computed.
    \item \textbf{Empty scene scan}: All detected objects are deactivated, and the scene is scanned again to produce the ground-truth empty scene point cloud.
    \item \textbf{Scan variations}: Additional scans are generated with configurable variations, including changes in furniture layout (by regenerating only the furniture while keeping the room structure), changes in scanner type (by switching to a different scanner preset), and changes in scanner position (by randomising the scanner location within the room volume).
\end{enumerate}

Each variation is stored alongside the main scan, together with its scan parameters, enabling controlled experiments on the impact of individual factors such as scanner type, scanner position, and temporal object changes. The variation system supports an arbitrary number of variation passes per room, with each pass independently configurable.

\subsection{The V-Scan Dataset}
\label{sec:3_vscan}

Using the framework described above, the V-Scan dataset was generated to serve as the primary evaluation benchmark for the scene and object completion stages of this thesis. The dataset is designed to provide the combination of properties that no existing dataset offers: scene-level partial scans paired with complete ground-truth geometry, multi-temporal and multi-sensory variations, and per-object ground truth including isolated partial scans, complete meshes, and oriented bounding boxes.

\subsubsection{Dataset Composition}

The V-Scan dataset contains 200 scene folders generated from 5 unique procedural room layouts (scene names), each scanned with multiple scanner type presets and variation passes. Table~\ref{tab:vscan_summary} summarises the key statistics of the dataset.

\begin{table}[h]
    \centering
    \caption[V-Scan dataset summary]{Summary statistics of the V-Scan dataset.}
    \label{tab:vscan_summary}
    \begin{tabular}{l r}
        \toprule
        \textbf{Metric} & \textbf{Value} \\
        \midrule
        Total scene folders & 200 \\
        Unique scene names & 5 \\
        Total objects & 3,100 \\
        Average objects per scene & 15.5 \\
        Scanner types used & 7 \\
        \bottomrule
    \end{tabular}
\end{table}

The point counts vary significantly across scanner types, reflecting the differences in scan density and range. Table~\ref{tab:vscan_points} provides representative point counts for a single scene scanned with different virtual scanner presets.

\begin{table}[h]
    \centering
    \caption[V-Scan representative point counts]{Representative point counts for a single scene (Industrial\_1) across different scanner types.}
    \label{tab:vscan_points}
    \begin{tabular}{l r}
        \toprule
        \textbf{Scanner Type} & \textbf{Points} \\
        \midrule
        Leica P30 & 2,529,458 \\
        Leica BLK360 Dense & $\sim$5,000,000 \\
        Leica BLK360 Fast & $\sim$1,500,000 \\
        Matterport Pro2 & $\sim$200,000 \\
        Matterport Pro3 & $\sim$400,000 \\
        iPhone LiDAR & $\sim$50,000 \\
        HoloLens 2 & $\sim$20,000 \\
        \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Per-Scene Outputs}

For each scene folder, the following outputs are available:

\begin{itemize}
    \item \textbf{Main scan}: coloured point cloud of the furnished scene with the main scanner type.
    \item \textbf{Panoramic image}: equirectangular panoramic image from the scanner position.
    \item \textbf{Scan parameters}: JSON file recording the scanner type configuration used.
    \item \textbf{Room parameters}: JSON file recording the procedural generation parameters.
    \item \textbf{Object bounding boxes}: JSON file containing oriented bounding boxes for all detected objects.
    \item \textbf{Per-object data}: for each object, the isolated partial point cloud, the ground-truth complete mesh (PLY), and the individual bounding box.
    \item \textbf{Empty scene scan}: coloured point cloud and panoramic image of the scene with all objects removed.
    \item \textbf{Scan variations}: additional point clouds and panoramic images for each configured variation, along with variation parameters.
\end{itemize}

This comprehensive set of outputs enables evaluation at multiple levels: scene-level completion quality can be assessed by comparing the reconstructed empty scene against the ground-truth empty scan, object-level completion quality can be assessed by comparing completed objects against the ground-truth meshes, and detection quality can be assessed by comparing detected bounding boxes against the ground-truth annotations.

% ─────────────────────────────────────────────────────────────────────────────
\section{Dataset Comparison and Overview}
\label{sec:3_comparison}

Table~\ref{tab:dataset_comparison} provides a comparative overview of all datasets used in this thesis. The comparison highlights the complementary nature of the data sources: real datasets provide realistic scanning characteristics but lack ground-truth completion data, while the synthetic V-Scan dataset provides comprehensive ground truth at the cost of reduced visual realism. By combining these sources, each pipeline stage can be evaluated using the most appropriate data.

\begin{table}[h]
    \centering 
    \caption[Dataset comparison]{Comparative overview of datasets used in this thesis. GT = ground truth available, MT = multi-temporal captures, MS = multiple scanner types.}
    \label{tab:dataset_comparison}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{l c c c c c c c}
        \toprule
        \textbf{Dataset} & \textbf{Type} & \textbf{GT Compl.} & \textbf{MT} & \textbf{MS} & \textbf{\# Scenes} & \textbf{Scene Types} & \textbf{Used For} \\
        \midrule
        ScanNet~\cite{dai_scannet_2017} & Real & \texttimes & \texttimes & \texttimes & 1,513 & Residential & Qual.\ eval. \\
        ScanNet++~\cite{yeshwanth_scannet_2023} & Real & \texttimes & \texttimes & \checkmark & 1,000+ & Mixed & Qual.\ eval. \\
        Matterport3D~\cite{chang_matterport3d_2017} & Real & \texttimes & \texttimes & \texttimes & 90 & Residential & Qual.\ eval. \\
        ARKitScenes~\cite{baruch_arkitscenes_2021} & Real & \texttimes & \texttimes & \texttimes & 5,047 & Mixed & Detection \\
        ShapeNet~\cite{chang_shapenet_2015} & Synthetic & \checkmark & -- & -- & -- & Objects only & Training \\
        Redwood~\cite{choi_large_2016} & Real & \checkmark & \texttimes & \texttimes & -- & Objects only & Training \\
        Campus captures & Real & \texttimes & \checkmark & \checkmark & 10+ & Industrial & Alignment \\
        V-Scan (ours) & Synthetic & \checkmark & \checkmark & \checkmark & 200 & Mixed & Quant.\ eval. \\
        \bottomrule
    \end{tabular}%
    }
\end{table}

The campus captures provide the multi-temporal and multi-sensory data required for the alignment experiments but lack ground-truth completion data. The public scene-level datasets provide large-scale, diverse real-world scenes for qualitative evaluation and generalisation assessment. ShapeNet and Redwood provide object-level data for training completion networks. The V-Scan dataset uniquely combines scene-level and object-level ground truth with multi-temporal and multi-sensory variations, enabling quantitative evaluation of the complete pipeline under cont 
rolled conditions.

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:3_conclusion}

This chapter presented the data acquisition strategy adopted in this thesis, addressing the diverse requirements of the proposed pipeline through a combination of three complementary data sources. Publicly available datasets provide large-scale, diverse real-world scenes for training, qualitative evaluation, and generalisation assessment. Real-world multi-temporal and multi-sensory captures from the university campus provide heterogeneous data for evaluating the alignment and updating pipeline. The novel V-Scan synthetic dataset, generated through a virtual scanning and procedural room generation framework, provides the comprehensive ground-truth data required for quantitative evaluation of scene and object completion methods.

The virtual scanning framework introduced in this chapter simulates the acquisition characteristics of seven different scanner types, ranging from high-end terrestrial laser scanners to consumer-grade XR devices, by configuring scan density, range, field of view, and noise parameters. The procedural room generator produces diverse furnished indoor environments with controllable complexity. Together, these components enable the automated generation of large-scale datasets with complete ground-truth geometry at both the scene and object level, multi-temporal variations through furniture regeneration, and multi-sensory variations through scanner type switching.

The V-Scan dataset, containing 200 scene folders with 3,100 total objects, serves as the primary benchmark for the quantitative evaluation conducted in the subsequent chapters. Its combination of scene-level and object-level ground truth, multi-temporal captures, and multi-sensory variations addresses the evaluation gap identified in Chapter~\ref{ch:introduction}, where the absence of ground truth for occluded regions was identified as a fundamental barrier to the quantitative assessment of completion methods.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\cleardoublepage